On the use of petri nets for modelling high-level programs

### 2010-07-19 Documents/Publications/type-2/type-2.tex
type-2: Towards a proof for Type-1 liveness and safety

  Despite the restrictions imposed by the production rules, the Type-2
  class of petri nets is powerful enough to represent programs written
  in traditional imperative languages such as C/C++ and also in
  synchronous languages such as Esterel. It has been used as a
  transition step in translating imperative programs to hardware
  descriptions\cite{ahir-thesis}. The resulting high-level synthesis
  flow uses the labelling scheme to optimise the generated hardware by
  identifying opportunities for sharing hardware resources in the
  absence of timing information.

@@

The structure of the Type-2 petri net is well suited for representing
the control flow observed in high-level programs --- in particular,
the interaction between concurrency and choice (branches). Threads
that are started within a branch must synchronise before exiting the
branch, and conversely, the targets of a branch encountered in a
thread must be restricted to that thread. A Type-2 petri net is built
using a set of construction rules that restrict the structure of the
petri net to ensure this behaviour. The same restrictions enable an
efficient labelling scheme that can be used to analyse the concurrency
of operations in the program. Note that the behaviour allowed by a
Type-2 petri net is a generalised version of a series-parallel
net\cite{something}, as can be seen in Section~\cite{sec:somewhere}.

@@

\begin{definition}
  A \emph{simple cycle} is the petri net $N = (P, T, E, M_0)$ where $P
  = \{p\}$, $T =\{t\}$, $E = \{(p,t), (t,p)\}$ and $M_0(p) = 1$.
\end{definition}

Clearly, a simple cycle is live and safe; and it is defined as the
simplest type-1 petri net. A type-1 petri net is defined as the petri
net obtained by repeatedly applying one more construction rules to the
simple cycle. Each rule replaces a simple place or transition in a
type-1 petri net with a petri net subgraph called a ``standard
region''. There are three types of standard regions, defined below.
The places occuring in a standard region are \textbf{not marked} in
the initial marking.

\begin{description}
\item [Series region:]
  A \emph{p-type series region} (t-type series region) is a petri net
  consiting of a sequence of simple transitions and places, where the
  first and the last elements are both places (transitions) with only
  one edge within the region. The first and the last elements
  are of the series region are called the \emph{entry} and the
  \emph{exit} of the region respectively.

@@

The construction rules of a Type-1 petri net can now be stated as the
following replacements:

\begin{srule}
  The petri net obtained by replacing a simple place (transition) in a
  Type-1 petri net by a p-type series region (t-type series region) is
  a also a Type-1 petri net.
\end{srule}

\begin{srule}
  The petri net obtained by replacing a simple place in a Type-1 petri
  net by a branch region is also a Type-1 petri net.
\end{srule}

@@

@@

\subsection{Series Region}
\label{sec:series-live-and-safe}

@@

A petri net formed by a branch region connected to simple marked place
is a \emph{state machine} which is strongly connected and has exactly
one token. Hence it is live and safe\ref{murata-tutorial}.

@@

All cycles in a petri net formed by a fork region connected to a
simple marked place $p_0$ pass through the place $p_0$. Thus every
cycle in the petri net has a marked place (hence it is live) and every
place is on a cycle with exactly one marked place (hence it is safe).

>>>>

  Despite the restrictions imposed by the production rules, the Type-2
  class of petri nets is powerful enough to represent programs written
  in traditional imperative languages such as C/C++ and also in
  synchronous languages such as Esterel. It has been used as a
  transition step in translating imperative programs to hardware
  descriptions\cite{ahir-thesis}. The resulting high-level synthesis
  flow uses the labelling scheme to identify opportunities for sharing
  hardware resources in the absence of timing information.

@@

The structure of the Type-2 petri net is well suited for representing
the control flow observed in high-level programs --- in particular,
the interaction between concurrency and choice (branches). Threads
that are started within a branch must synchronise before exiting the
branch, and conversely, the targets of a branch that occurs in a
thread must be restricted to that thread. A Type-2 petri net is built
using a set of construction rules that restrict the structure of the
petri net to ensure this behaviour. The same restrictions enable an
efficient labelling scheme that can be used to analyse the concurrency
of operations in the program. Note that the behaviour allowed by a
Type-2 petri net is a generalised version of a series-parallel
net\cite{something}, as can be seen in Section~\cite{sec:somewhere}.

@@

\begin{definition}
  A \emph{simple cycle} is the petri net $N = (P, T, E, M_0)$ where $P
  = \{p\}$, $T =\{t\}$, $E = \{(p,t), (t,p)\}$ and the
  initial marking $M_0(p) = 1$.
\end{definition}

Clearly, a simple cycle is live and safe. It is defined as the
simplest type-1 petri net, and any petri net obtained by repeatedly
applying one or more construction rules to the simple cycle is also a
Type-1 petri net. Each rule replaces a simple place or transition in
the original petri net with a subgraph called a ``standard region''.
There are three types of standard regions, defined below. Places
occuring in the standard regions are \textbf{not marked} in the
initial marking.

\begin{description}
\item [Series region:] A \emph{p-type series region} (t-type series
  region) is a petri net consiting of a sequence of simple transitions
  and places, where the first and the last elements are both places
  (transitions) with only one edge within the region. The first and
  the last elements of the series region are called the \emph{entry}
  and the \emph{exit} of the region respectively.

@@

The construction rules for obtaining a  Type-1 petri net can now be
stated as follows:

\setcounter{srule}{-1}

\begin{srule}
  The simple cycle is a Type-1 petri net.
\end{srule}

\begin{srule}
  The petri net obtained by replacing an unmarked simple place
  (transition) in a Type-1 petri net by a p-type series region (t-type
  series region) is a also a Type-1 petri net.
\end{srule}

\begin{srule}
  The petri net obtained by replacing an unmarked simple place in a
  Type-1 petri net by a branch region is also a Type-1 petri net.
\end{srule}

@@

The above construction rules ensure that the place occuring in the
original simple cycle is the only marked place and it is never
replaced during the construction of the Type-1 petri net.

@@

\subsection{Simplification}
\label{sec:simplification-live-safe}

Consider a simple place $p$ in a live and safe petri net $N =
(P,T,E,M)$. Let $u,v$ be the predecessor and successor transitions of
$p$ in $N$, i.e, $u,v \in T$ and $(u,p), (p,v)\in E$. Since $N$ is
live and safe, it follows that when $u$ fires, $v$ must fire before
$u$ may fire again (safety), and $v$ must eventually fire when $u$
fires (liveness). Thus, any firing sequence of the petri net $N$ that
includes the firing of $u$ (and consequently also $v$) effectively
puts a token in $p$ and then eventually removes it. Thus, the petri
net $N' = N - \{p,u,v\}$ can be modelled as an initially marked simple
place $p_0$ that forms a cycle with $u$, $v$ and $p$. Similarly,
relative to a simple transition $t$, the rest of the petri net can be
modelled as an initially marked simple place $p_0$ that forms a simple
cycle with $t$. Both replacements are valid if and only if the
original petri net is live and safe; the resulting simple cycle is
also live and safe. We use this simplification to examine the effect
of applying a construction rule on a Type-1 petri net:

\begin{enumerate}
\item When a rule replaces a simple place $p$, the rest of the petri
  net is reduced to a simple marked place $p_0$ connected to $p$ by
  simple transitions.
\item When a rule replaces a simple transition $t$, the rest of the
  petri net is first reduced to a single marked place $p_0$.
\end{enumerate}

Thus, in each of the following subsections, we model an arbitrary
Type-1 petri-net as a cycle consisting of a marked place $p_0$
connected to the candidate simple place or transition through a
suitable number of additional simple transitions. We refer to this
cycle as the \emph{base petri net} in the rest of this section.

\subsection{p-type Series Region}
\label{sec:series-live-and-safe}

When a simple place (transition) in the base petri net is replaced by
p-type series region (t-type series region), the result is simply a
larger cycle. The resulting petri net is live and safe, since it is a
strongly connected \emph{state machine} with exactly one marked place.

@@

When a place in the base petri net is replaced by a branch region, the
resulting petri net is a strongly connected \emph{state machine} which
is that has exactly one marked place. Hence it is live and
safe\ref{murata-tutorial}.

@@

When a transition in the base petri net is replaced by a fork region,
the resulting petri net is a \emph{marked graph} such that every cycle
in the marked graph pass through the single marked place $p_0$. Thus
every cycle in the petri net has a marked place (hence it is live) and
every place is on a cycle with exactly one marked place (hence it is
safe).
